WWTF Grant Proposal\\Leveraging Resolution for Verified Program Synthesis

### 2022-03-21 WWTF/Draft Alexander/Proposal.tex
Initial Draft Idea

	Program synthesis is a long-standing problem crossing many areas of computer science and mathematics including programming languages, artificial intelligence, formal methods and proof theory. There have been major advances in NLP (natural language processing) based program synthesis recently, most prominently DeeepMind's AlphaCode~\cite{AlphaCode1}, OpenAI's Codex~\cite{Codex1} and Github's Copilot~\cite{Copilot1}.

@@

	Another route to program synthesis is offered through the Curry-Howard correspondence. Exploiting the equivalence of proofs in intuitionistic logic and functional programs allows us to reduce the problem of synthesising a program satisfying a specification into proving its validity. The added benefit of this approach is that a program generated from such a proof is guaranteed to be correct. From the perspective of formal methods there are benefits as well. At the moment creating a verified program requires both writing specification and code. Being able to automate the latter step would save considerable overhead.

	However automated theorem proving in intuitionistic logic is notoriously difficult in particular due to the absence of convenient normal forms. Most state of the art theorem provers like Vampire~\cite{Vampire1} and E~\cite{E1} use resolution which is a great technique for automation but inherently non-intuitionistic in that it requires the formulas to be transformed into conjunctive normal form and works through refutation of the negated formula, which is not  intuitionistically sufficient for a proof of validity.

@@

	Program synthesis in the sense of this project is concerned with generating a program that satisfies a specification given in FOL (first order logic). Instead of directly generating a program however we obtain it via the Curry-Howard correspondence from a proof of the validity of the specification. Such a program will always be valid.

@@

	However for the Curry-Howard correspondence to apply the initial proof has to be valid in intuitionistic logic but state-of-the-art theorem provers utilize resolution which is an inherently classical techniques. There are 2 approaches  present in literature that can help to remedy this fact.

@@

	On the other hand in certain regimes it is possible to transform a classical proof into an intuitionistic one, e.g. proofs of $\Pi_2$ sentences in Peano Arithmetic~\cite{HAPA}, via Friedman's Translation or similar methods.~\cite{Friedman} This approach has been examined in different small-scale settings and would have to be adapted for a general setting.

@@

	\bibliographystyle{acm}
	\bibliography{references}

>>>>

	Program synthesis is a long-standing problem crossing many areas of computer science and mathematics including programming languages, artificial intelligence, formal methods and proof theory. There have been major advances in NLP (natural language processing) based program synthesis recently, most prominently DeeepMind's AlphaCode, OpenAI's Codex and Github's Copilot.

@@

	Another route to program synthesis is offered through the Curry-Howard correspondence. Exploiting the equivalence of proofs in intuitionistic logic and functional programs allows us to reduce the problem of synthesising a program satisfying a specification into proving the validity of said specification. The added benefit of this approach is that a program generated from such a proof is guaranteed to be correct.

	However automated theorem proving in intuitionistic logic is notoriously difficult in particular due to the absence of convenient normal forms. Most state of the art provers use techniques which are inherently non-intuitionistic.

@@

	Program synthesis in the sense of this project is concerned with generating a program that satisfies a specification given in FOL (first order logic). Instead of directly generating a program however we obtain it via the Curry-Howard correspondence from a proof of the validity of the specification. Such a program will always satisfy the specification.

@@

	However for the Curry-Howard correspondence to apply the initial proof has to be valid in intuitionistic logic but state-of-the-art theorem provers utilize resolution which is an inherently classical technique. There are 2 approaches  present in literature that can help to remedy this.

@@

	On the other hand in certain regimes it is possible to transform a classical proof into an intuitionistic one, e.g. proofs of $\Pi_2$ sentences in Peano Arithmetic~\cite{HAPA}, via Friedman's Translation or similar methods.~\cite{Friedman} This approach has been examined in different model settings and would have to be adapted and extended for our general problem.

@@

	\bibliographystyle{plain}
	\bibliography{references}
